---
title: "WGBS Analysis Pipeline"
author: "Mark E. Pepin"
date: "05/05/2018"
output:
  pdf_document:
    toc: yes
  html_document:
    code_folding: hide
    keep_md: yes
    toc: yes
    toc_float: yes
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage[table]{xcolor}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
mainfont: Times
fontsize: 10pt
always_allow_html: yes
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(tidy.opts=list(width.cutoff=30),tidy=FALSE, warning = FALSE, message = FALSE, cache = TRUE, cache.lazy = FALSE)
options(knitr.kable.NA = '')
```

**Author**: Mark E. Pepin, MS Biomedical Engineering | MD/PhD Trainee  
**Contact**: pepinme@gmail.com  
**Institution**: University of Alabama at Birmingham  
**Location**: 542 Biomedical Research Building 2, Birmingham, AL 35294  

#Genome Assembly and Alignment

The first task is to align the bisulfite reduced and sequenced reads to a genome assembly. To accomplish this, I prepared the genome assembly based on Gencode annotation (gencode.v28.annotation.gtf) and sequence (GRCh38.p12.genome.fa). For whole-genome bisulfite sequencing via the Bismark (v0.20.0) aligner and genome preparation, a CT- and GA-converted assemblies are created.

##Genome Assembly

`./bismark_genome_preparation --path_to_bowtie ../bowtie2-2.3.4.2-linux-x86_64 -- verbose ../../Input/Genome/GRCh38.p12.genome.fa`

##Adapter Trimming

Once the genome assembly was created, adapter sequences were trimmed and sequencing quality assessed via trim_galore and FastQC, respectively.

`module load SAMtools/1.6-intel-2017a`
`module load Bowtie2/2.3.3-intel-2017a`
`module load Trim_Galore/0.4.4-foss-2016b`
`module load FastQC/0.11.7-Java-1.8.0_74`

`trim_galore -o $INPUT_DIR/fastq_trimmed/ --paired --rrbs --non_directional --length 20 --fastqc` `$INPUT_DIR/fastq/${VAR}_1.txt.gz $INPUT_DIR/fastq/${VAR}_2.txt.gz`

##Read Alignment

We then aligned all 34 paired-end .fastq files to the genome assemblies using the following command:

`$BISMARK/bismark \`
`--bowtie2 --bam $GENOME_DIR \`
`-1 $INPUT_DIR/fastq_trimmed/${VAR}_1.txt.gz_val_1.fq.gz -2 $INPUT_DIR/fastq_trimmed/${VAR}_2.txt.gz_val_2.fq.gz \`
`--output_dir $RESULTS_DIR/WGBS`

##Deduplication

Once aligned, we need to "deduplicate" the aligned .bam files to reduce PCR bias.

`$BISMARK/deduplicate_bismark \`
`--output_dir $RESULTS_DIR/WGBS/deduplicate_bismark \`
`--bam -p \`
`$RESULTS_DIR/WGBS/${VAR}_1.txt.gz_val_1_bismark_bt2_pe.bam`

##Methylation Extraction

Once finished, the CpG methylation was extracted as both bedgraph file (for UCSC genome browser) and bed file, which was then used to identify differentially-methylated cytosines (DMCs) and differentially-methylated regions (DMRs).

`$BISMARK/bismark_methylation_extractor \`
`-p --no_overlap --report --bedGraph --gzip \`
`$RESULTS_DIR/WGBS/deduplicate_bismark/${VAR}_1.txt.gz_val_1_bismark_bt2_pe.deduplicated.bam`

The "bismark.cov" files that resulted from this were then read into R () and combined into a single "object" for differential methylation analysis

#Differential Methylation Analysis

##Combining sample methylation

```{r Count.Compile}
#Conditions to be used in differential methylation analysis (FILL OUT)
library(openxlsx)
TREATMENT=c("VEH","AZA")
CELL=c("ASC")

ANALYSIS="ASC_AZA.v.VEH"
### "2" is Pre-LVAD, "3" is Post-LVAD, "1" is CON
library(methylKit)
file.list <- list.files(path = paste0("../1_Input/Methyl/2_bwameth.out/", CELL), pattern = "*.counted_CpG.bedGraph", full.names = TRUE, all.files = TRUE)
#Generate Column names (remove the extra nonsense from the path names)
colnames <- gsub( "*.counted_CpG.bedGraph", "", file.list)
colnames <- gsub( paste0("[.][.]/1_Input/Methyl/2_bwameth.out/", CELL, "/"), "", colnames)
sample_id<-as.list(colnames)
#Import the Index file
Index.raw<-read.xlsx("../1_Input/Index/Index_Napoli.xlsx", sheet = CELL)
# Index.raw$Treatment<-factor(Index.raw$Group, levels = c("CON", "LOW", "HIGH"))

## Sort the index according to the .bed file ordering (as imported). This is important for correct annotation of samples.
Index_sorted<-subset(Index.raw, Sample.ID %in% colnames)
Index_sorted$Sample.ID<-factor(Index_sorted$Sample.ID, levels = colnames)
Index_sorted<-Index_sorted[order(Index_sorted$Sample.ID),]
#Created an index based on the desired sample criteria (Phenotype and Timing)
Index_filtered<-subset(Index_sorted, Treatment %in% TREATMENT)
##Change "Pre/Post" variable to a factor of "0" or "1" (needed for differential methylation)
Index_sorted$Treatment<-factor(Index_filtered$Treatment, levels = c("VEH", "AZA"))
Index_sorted$Treatment<-as.integer(Index_sorted$Treatment)
Index_sorted$Treatment<-c(1,1,0,0)

# Index_sorted<-Index_sorted[!is.na(Index_sorted$Timing),]
##Create a methlRawlistDB
file.list<-as.list(file.list)
myobj<-methRead(file.list, sample.id = sample_id, assembly = "hg38", treatment = Index_sorted$Treatment, pipeline = "bismarkCoverage", header = FALSE, context = "CpG")
##Example of smaple statistics (can spot check these)
getMethylationStats(myobj[[2]], plot = F, both.strands = F)

#Subset the methylRawList to include only the sample_id's for the desired analysis
myobj_filtered<-reorganize(myobj, sample.ids = Index_filtered$Sample.ID, Index_sorted$Treatment)
```

Once the samples have been compiled, it is valuable to perform some basic visualizations and statistics to determine whether quality filtering is necessary. The distribution of methylation change is plotted as a histogram (typically bimodal at the extremes), as well as a distribution of the read coverage per based, again plotted as a histogram. For the latter plot, it is important to determine whether PCR duplication biases the read coverage. If so, a secondary peak would emerge on the right-most portion of the histogram. In the current analysis, coverage distribution exhibits a one-tailed distribution, suggesting that the "deduplication" step in the alignment effectively eliminated the PCR amplification bias in coverage.

```{r Methylation_stats}
library(graphics)
getMethylationStats(myobj_filtered[[2]], plot = T, both.strands = F)
getCoverageStats(myobj_filtered[[2]], plot = T, both.strands = F)
#Save these files in an output folder
ifelse(!dir.exists(file.path("../2_Output/", ANALYSIS)), dir.create(file.path("../2_Output/", ANALYSIS)), FALSE)
pdf(file=paste0("../2_Output/", ANALYSIS, "/", ANALYSIS, "_Methylation.Stats.pdf"))
getMethylationStats(myobj_filtered[[2]], plot = T, both.strands = F)
dev.off()
pdf(file=paste0("../2_Output/", ANALYSIS, "/", ANALYSIS, "_Coverage.Stats.pdf"))
getCoverageStats(myobj_filtered[[2]], plot = T, both.strands = F)
dev.off()
```

Although most important in the context of correcting PCR-bias (duplication), filtering samples based on coverage also reduces false discovery based on low-coverage genomic regions. If PCR bias exists, an artificially high coverage would exist. Low coverage is also a concern due to low statistical power associated with low-coverage regions. Below, we discard bases with coverage below 10X, but also discard bases with coverage > 99.9th percentile.

```{r filter_coverage}
#remove exceedingly high-coverage (risk of PCR bias) or low-coverage DMPs (low statistical power) 
filtered.myobj <- filterByCoverage(myobj_filtered, lo.count = 5, lo.perc = NULL, hi.count = NULL, hi.perc = 99.9)
```

##Visualizing Methylation

```{r Methylation_visualization}
#destrand and unite the sample data
meth<-unite(filtered.myobj, destrand = FALSE) #When calculating DMRs, it is not helpful to "destrand"
clusterSamples(meth, dist = "correlation", method = "ward", plot = TRUE)
PCASamples(meth, screeplot = TRUE)
PCASamples(meth)
#Create a folder in which to generate all documents/tables for this analysyis
ifelse(!dir.exists(file.path("../2_Output/", ANALYSIS)), dir.create(file.path("../2_Output/", ANALYSIS)), FALSE)
#Create dendrogram and PCA plots
pdf(file=paste0("../2_Output/", ANALYSIS, "/", ANALYSIS, "_Clustering.pdf"))
clusterSamples(meth, dist = "correlation", method = "ward", plot = TRUE)
PCASamples(meth, screeplot = TRUE)
PCASamples(meth)
dev.off()
```


##Tiling Methylation Windows

```{r tiling}
library(dplyr)
##################################
#Differential Methylation of Tiles
##################################
tiles = tileMethylCounts(myobj_filtered, win.size = 500, step.size = 500)
meth_tile<-unite(tiles, destrand = FALSE) #When calculating DMRs, it is not helpful to "destrand"
myDiff_tiles=calculateDiffMeth(meth_tile, test = "F", mc.cores = 7)
myDiff_tile.md<-as(myDiff_tiles,"methylDiff")
myDiff_tiles.filtered<-dplyr::select(myDiff_tile.md, chr, start, end, strand, meth.diff, pvalue, qvalue)

#Check clustering of samples by DMR correlation
clusterSamples(meth_tile, dist = "correlation", method = "ward", plot = TRUE)
PCASamples(meth_tile, screeplot = TRUE)
PCASamples(meth_tile)

#Create a folder in which to generate all documents/tables for this analysyis
ifelse(!dir.exists(file.path("../2_Output/", ANALYSIS, "DMR/")), dir.create(file.path("../2_Output/", ANALYSIS, "DMR/")), FALSE)
#Create dendrogram and PCA plots
pdf(file=paste0("../2_Output/", ANALYSIS, "/DMR/", ANALYSIS, "_Clustering.pdf"))
clusterSamples(meth_tile, dist = "correlation", method = "ward", plot = TRUE)
PCASamples(meth, screeplot = TRUE)
PCASamples(meth)
dev.off()

##############################################
#Calculate percent methylation for each sample/site
##############################################

Methylation<-as.data.frame(meth_tile)
f = function(Cyt, cov, col_name) {
  require(lazyeval)
  require(dplyr)
    mutate_call = lazyeval::interp(~ (a / b)*100, a = as.name(Cyt), b = as.name(cov))
    Methylation %>% mutate_(.dots = setNames(list(mutate_call), col_name))
}
for(i in seq_along(Index_filtered$Sample.ID)){
  COVERAGE=paste0("coverage", i)
  mC=paste0("numCs", i)
  perc.mC=paste0("perc.mC_", Index_filtered$Sample.ID[i])
  Methylation<-f(Cyt=mC, cov=COVERAGE, col_name=perc.mC)
}
Methylation<-dplyr::select(Methylation, chr, start, end, contains("perc.mC"))

#Merge with the percent methylation (by cytosine)
myDiff_tiles.filtered<-left_join(myDiff_tiles.filtered, Methylation)

#Subset by statistical threshold
myDiff.tiles_p05<-dplyr::filter(myDiff_tiles.filtered, pvalue<0.05)
myDiff.tiles_q05<-dplyr::filter(myDiff_tiles.filtered, qvalue<0.05)

#Save a copy of the differential Methylation analysis
wb_countData<-createWorkbook()
addWorksheet(wb_countData, "P < 0.05")
  writeData(wb_countData, "P < 0.05", myDiff.tiles_p05, rowNames = F)
addWorksheet(wb_countData, "Q < 0.05")
  writeData(wb_countData, "Q < 0.05", myDiff.tiles_q05, rowNames = F)
saveWorkbook(wb_countData, file = paste0("../2_Output/", ANALYSIS, "/DMR/", ANALYSIS, "_DiffMeth.xlsx"), overwrite = TRUE)

myDiff.files_q05_GR<-makeGRangesFromDataFrame(myDiff.tiles_q05, seqnames.field = "chr", strand.field="strand", start.field = "start", end.field = "end", keep.extra.columns = T)

write.table(myDiff.tiles_q05, file = "../2_Output/Tiles_Q05_DiffMeth.bed", sep = "\t", row.names = F, col.names = F, quote = F)

############################
##ANNOTATION
############################
library(openxlsx)
library(annotatr)
library(AnnotationHub)
library(rtracklayer)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
#convert to GRanges object (correct format for annotatr)
myDiff.tiles_p05_GR<-makeGRangesFromDataFrame(myDiff.tiles_p05, seqnames.field = "chr", strand.field="strand", start.field = "start", end.field = "end", keep.extra.columns = T)
#create annotations from the following sources
annots = c('hg38_cpgs', 'hg38_basicgenes')
# Build the annotations (a single GRanges object)
annotations = build_annotations(genome = 'hg38', annotations = annots)
# myDiff_GR<-as(myDiff, "GRanges")
# Intersect the regions read in with the annotations
dm_annotated.tiles = annotate_regions(
    regions = myDiff.tiles_p05_GR,
    annotations = annotations,
    ignore.strand = TRUE,
    quiet = FALSE)
#convert to a data.frame
df_dm_annotated.tiles = data.frame(dm_annotated.tiles)
# A GRanges object is returned
print(dm_annotated.tiles)
##The issue with this annotation is that each DMP has multiple repeated rows if different annotations. To simplify this, we can condense the annotations into strings. This makes the resulting file more manageable based on the differential-methylation data.
DiffMeth_Annotated.tiles<-df_dm_annotated.tiles %>% 
  tidyr::fill(annot.symbol) %>% distinct() %>%
  dplyr::group_by(seqnames, start, end, meth.diff, pvalue, qvalue, annot.symbol) %>% 
  dplyr::summarise(Annotation=paste(unique(annot.type), collapse = ";"), Test=paste(unique(annot.id), collapse = ";"))
#Add %Methylation
DiffMeth_Annotated.tiles<-dplyr::rename(DiffMeth_Annotated.tiles, chr=seqnames)
DiffMeth_Annotated.tiles<-dplyr::left_join(DiffMeth_Annotated.tiles, Methylation)
#subset the Differential Methylation by statistics
DiffMeth_Annotated.tiles_p05<-subset(DiffMeth_Annotated.tiles, pvalue<0.05)
DiffMeth_Annotated.tiles_q05<-subset(DiffMeth_Annotated.tiles, qvalue<0.05)
#Write out the annotated DMP file 
library(openxlsx)
ifelse(!dir.exists(file.path("../2_Output/", ANALYSIS, "/DMR/")), dir.create(file.path("../2_Output/", ANALYSIS, "/DMR/")), FALSE)
wb_WGBS_Annotate<-createWorkbook()
addWorksheet(wb_WGBS_Annotate, "P < 0.05")
  writeData(wb_WGBS_Annotate, "P < 0.05", DiffMeth_Annotated.tiles_p05, rowNames = F)
addWorksheet(wb_WGBS_Annotate, "Q < 0.05")
  writeData(wb_WGBS_Annotate, "Q < 0.05", DiffMeth_Annotated.tiles_q05, rowNames = F)
saveWorkbook(wb_WGBS_Annotate, file = paste0("../2_Output/", ANALYSIS, "/DMR/", ANALYSIS, "_Annotated_DiffMeth.xlsx"), overwrite = TRUE)
#Provide a summary of the annotation
dm_annsum.tile = summarize_annotations(
    annotated_regions = dm_annotated.tiles,
    quiet = TRUE)
print(dm_annsum.tile)
#Plot the annotation distribution
dm_vs_kg_annotations.tile = plot_annotation(
    annotated_regions = dm_annotated.tiles,
    plot_title = '# of Sites Tested for DM annotated on chr9',
    x_label = 'knownGene Annotations',
    y_label = 'Count')
print(dm_vs_kg_annotations.tile)

annots_order = c(
    'hg38_genes_1to5kb',
    'hg38_genes_promoters',
    'hg38_genes_5UTRs',
    'hg38_genes_exons',
    'hg38_genes_introns',
    'hg38_genes_3UTRs')
dm_vs_kg_annotations = plot_annotation(
    annotated_regions = dm_annotated.tiles,
    annotation_order = annots_order,
    plot_title = '# of Sites Tested for DM annotated on chr9',
    x_label = 'knownGene Annotations',
    y_label = 'Count')
print(dm_vs_kg_annotations)
```

## Differentially-Methylated Cytosines (DMCs)

```{r DMC}
#Calculate the differential methylation (TAKES A LONG TIME!)
library(openxlsx)
myDiff<-calculateDiffMeth(meth, test = "F", mc.cores = 7)
myDiff_md<-as(myDiff,"methylDiff")
myDiff_filtered<-dplyr::select(myDiff_md, chr, start, end, strand, meth.diff, pvalue, qvalue)

#Calculate percent methylation for each sample/site
Methylation<-as.data.frame(meth)
f = function(Cyt, cov, col_name) {
  require(lazyeval)
  require(dplyr)
    mutate_call = lazyeval::interp(~ (a / b)*100, a = as.name(Cyt), b = as.name(cov))
    Methylation %>% mutate_(.dots = setNames(list(mutate_call), col_name))
}
for(i in seq_along(Index_filtered$Sample.ID)){
  COVERAGE=paste0("coverage", i)
  mC=paste0("numCs", i)
  perc.mC=paste0("perc.mC_", Index_filtered$Sample.ID[i])
  Methylation<-f(Cyt=mC, cov=COVERAGE, col_name=perc.mC)
}
Methylation<-dplyr::select(Methylation, chr, start, end, contains("perc.mC"))

#Merge with the percent methylation (by cytosine)
myDiff_filtered<-left_join(myDiff_filtered, Methylation)

#Subset by statistical threshold
myDiff_p05<-dplyr::filter(myDiff_filtered, pvalue<0.05)
myDiff_q05<-dplyr::filter(myDiff_filtered, qvalue<0.05)

#Save a copy of the differential Methylation analysis
wb_countData<-createWorkbook()
addWorksheet(wb_countData, "P < 0.05")
  writeData(wb_countData, "P < 0.05", myDiff_p05, rowNames = F)
addWorksheet(wb_countData, "Q < 0.05")
  writeData(wb_countData, "Q < 0.05", myDiff_q05, rowNames = F)
saveWorkbook(wb_countData, file = paste0("../2_Output/", ANALYSIS, "/", ANALYSIS, "_DiffMeth.xlsx"), overwrite = TRUE)
write.table(myDiff_p05, file=paste0("../2_Output/", ANALYSIS, "/", ANALYSIS, "_DiffMeth.bed"), quote=F, sep="\t", row.names=F, col.names=F)
```

##Annotate DMPs with Genomic and CpG Loci

```{r Annotate_DMP}
library(openxlsx)
library(annotatr)
library(AnnotationHub)
library(rtracklayer)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)

#convert to GRanges object (correct format for annotatr)
myDiff_p05_GR<-makeGRangesFromDataFrame(myDiff_p05, seqnames.field = "chr", strand.field="strand", start.field = "start", end.field = "end", keep.extra.columns = T)
#create annotations from the following sources
annots = c('hg38_cpgs', 'hg38_basicgenes')
# Build the annotations (a single GRanges object)
annotations = build_annotations(genome = 'hg38', annotations = annots)
# myDiff_GR<-as(myDiff, "GRanges")
# Intersect the regions read in with the annotations
dm_annotated = annotate_regions(
    regions = myDiff_p05_GR,
    annotations = annotations,
    ignore.strand = TRUE,
    quiet = FALSE)
#convert to a data.frame
df_dm_annotated = data.frame(dm_annotated)
# A GRanges object is returned
print(dm_annotated)
##The issue with this annotation is that each DMP has multiple repeated rows if different annotations. To simplify this, we can condense the annotations into strings. This makes the resulting file more manageable based on the differential-methylation data.
DiffMeth_Annotated<-df_dm_annotated %>% 
  tidyr::fill(annot.symbol) %>% distinct() %>%
  dplyr::group_by(seqnames, start, end, meth.diff, pvalue, qvalue, annot.symbol) %>% 
  dplyr::summarise(Annotation=paste(unique(annot.type), collapse = ";"), Test=paste(unique(annot.id), collapse = ";"))
#Add %Methylation
DiffMeth_Annotated<-dplyr::rename(DiffMeth_Annotated, chr=seqnames)
DiffMeth_Annotated<-dplyr::left_join(DiffMeth_Annotated, Methylation)
#subset the Differential Methylation by statistics
DiffMeth_Annotated_p05<-subset(DiffMeth_Annotated, pvalue<0.05)
DiffMeth_Annotated_q05<-subset(DiffMeth_Annotated, qvalue<0.05)
#Write out the annotated DMP file 
library(openxlsx)
ifelse(!dir.exists(file.path("../2_Output/", ANALYSIS)), dir.create(file.path("../2_Output/", ANALYSIS)), FALSE)
wb_WGBS_Annotate<-createWorkbook()
addWorksheet(wb_WGBS_Annotate, "P < 0.05")
  writeData(wb_WGBS_Annotate, "P < 0.05", DiffMeth_Annotated_p05, rowNames = F)
addWorksheet(wb_WGBS_Annotate, "Q < 0.05")
  writeData(wb_WGBS_Annotate, "Q < 0.05", DiffMeth_Annotated_q05, rowNames = F)
saveWorkbook(wb_WGBS_Annotate, file = paste0("../2_Output/", ANALYSIS, "/", ANALYSIS, "_Annotated_DiffMeth.xlsx"), overwrite = TRUE)
#Provide a summary of the annotation
dm_annsum = summarize_annotations(
    annotated_regions = dm_annotated,
    quiet = TRUE)
print(dm_annsum)
#Plot the annotation distribution
dm_vs_kg_annotations = plot_annotation(
    annotated_regions = dm_annotated,
    plot_title = '# of Sites Tested for DM annotated on chr9',
    x_label = 'knownGene Annotations',
    y_label = 'Count')
print(dm_vs_kg_annotations)

annots_order = c(
    'hg38_genes_1to5kb',
    'hg38_genes_promoters',
    'hg38_genes_5UTRs',
    'hg38_genes_exons',
    'hg38_genes_introns',
    'hg38_genes_3UTRs')
dm_vs_kg_annotations = plot_annotation(
    annotated_regions = dm_annotated,
    annotation_order = annots_order,
    plot_title = '# of Sites Tested for DM annotated on chr9',
    x_label = 'knownGene Annotations',
    y_label = 'Count')
print(dm_vs_kg_annotations)
```

#Heatmap of Differential Methylation

```{r Heatmap_DMPs}
library(pheatmap)
hm_Data<-as.data.frame(DiffMeth_Annotated_p05)
hm_Data<-hm_Data[!is.na(hm_Data$annot.symbol),]
rownames(hm_Data)<-make.unique(hm_Data$annot.symbol, sep = ".")

##Make heatmap
STATISTIC=0.05
hm_Data<-dplyr::filter(hm_Data, pvalue<STATISTIC)

hm_Data<-dplyr::select(myDiff.tiles_p05, contains("perc.mC"))
hm_Data<-data.matrix(hm_Data)

##
##Index file for annotating samples
hm_Index<-Index_filtered
hm_Index$Sample.ID<-paste0("perc.mC_", hm_Index$Sample.ID)
rownames(hm_Index)<-hm_Index$Sample.ID
hm_Index<-as.data.frame(hm_Index)
hm_Index<-dplyr::select(hm_Index, Sample.ID, Treatment)

paletteLength <- 100
myColor <- colorRampPalette(c("dodgerblue4", "white", "gold2"))(paletteLength)
pheatmap(hm_Data,
         cluster_cols=T, 
         border_color=NA, 
         cluster_rows=T, 
         scale = 'row',
         show_colnames = T, 
         show_rownames = F, 
         color = myColor,
         annotation_col = hm_Index, 
         filename = paste0("../2_Output/", ANALYSIS, "/DMR/", ANALYSIS, "_Heatmap.Q05.pdf"))
```


#Methylation Distribution using **EnrichedHeatmap**

```{r Enriched.Heatmap}
#Import the genomic annotation file
library(EnrichedHeatmap)
library(annotatr)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
annots = c("hg38_basicgenes", "hg38_genes_promoters", "hg38_genes_intergenic",
           "hg38_genes_intronexonboundaries", "hg38_cpgs", "hg38_cpg_islands", "hg38_cpg_shores", "hg38_cpg_shelves", "hg38_cpg_inter")
annotations=build_annotations(genome = "hg38", annotations = annots)
annotations<-keepStandardChromosomes(annotations, pruning.mode = "coarse") #Remove nonstandard chromosomes
myDiff_p05<-dplyr::mutate(as.data.frame(myDiff_p05), absolute.meth=abs(meth.diff))
myDiff_p05<-as(myDiff_p05, "GRanges")
meDiff<-as(myDiff, "GRanges")

# #Import the annotated "target" data
# myDiff<-openxlsx::read.xlsx(paste0("../2_Input/WGBS_MethylKit_DiffMeth", ANALYSIS,".xlsx"), sheet = "P < 0.05")
# myDiff_GR<-as(myDiff, "GRanges")

#Annotate GRanges using hg38 genome
dm_annotated = annotate_regions(
  regions = myDiff_p05,
  annotations = annotations,
  ignore.strand = TRUE)
#create data.frame
df_dm_annotated <- as.data.frame(dm_annotated)
#
library(GenomicFeatures)
genes<-genes(TxDb.Hsapiens.UCSC.hg38.knownGene)
tss = promoters(genes, upstream = 0, downstream = 1)
mat1 = normalizeToMatrix(myDiff_p05, tss, value_column = "absolute.meth", extend = 5000, mean_mode="absolute", w=50, keep = c(0, 0.99))
EnrichedHeatmap(mat1)

pdf(file = paste0("../2_Output/", ANALYSIS, "/","_Methyl.Gene.Distribution.pdf"), height = 5, width = 5)
EnrichedHeatmap(mat1)
dev.off()

partition = kmeans(mat1, centers = 3)$cluster
lgd = Legend(at = c("cluster1", "cluster2", "cluster3"), title = "Clusters", 
    type = "lines", legend_gp = gpar(col = 2:4))
ht_list = Heatmap(partition, col = structure(2:4, names = as.character(1:3)), name = "partition",
              show_row_names = FALSE, width = unit(3, "mm")) + EnrichedHeatmap(mat1, col = c("white", "red"), name = "% Methylation - Heart Failure", split = partition, top_annotation = HeatmapAnnotation(lines = anno_enriched(gp = gpar(col = 2:4))), column_title = "|PercentMethylation|")
draw(ht_list, main_heatmap = "% Methylation - Heart Failure")

partition = kmeans(mat1, centers = 3)$cluster
lgd = Legend(at = c("cluster1", "cluster2", "cluster3"), title = "Clusters", 
    type = "lines", legend_gp = gpar(col = 2:4))
ht_list = Heatmap(partition, col = structure(2:4, names = as.character(1:3)), name = "partition",
              show_row_names = FALSE, width = unit(3, "mm")) + EnrichedHeatmap(mat1, col = c("white", "red"), name = "% Methylation - Heart Failure", split = partition, top_annotation = HeatmapAnnotation(lines = anno_enriched(gp = gpar(col = 2:4))), column_title = "|PercentMethylation|")
pdf(file = paste0("../2_Output/", ANALYSIS, "/", ANALYSIS, "_Methyl.Gene.Distribution_Kmeans.pdf"), height = 7, width = 5)
draw(ht_list, main_heatmap = "% Methylation - Heart Failure")
dev.off()
```

#Candidate Genes

```{r Candidate DMPs}
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(dplyr)
library(gtools)
######################################################
VAR1="Response"
VAR2="Timing"
GENES=c("HDAC2")
RESPO=c("CON", "NR")
TIM=c("CON","Pre")
PAIRS=c("CON" ,"Paired")

ifelse(!dir.exists(file.path(paste0("../2_Output/2_Methyl/Candidates/"))), dir.create(file.path(paste0("../2_Output/2_Methyl/Candidates/"))), FALSE)

######################################################
#Import dataset
Counts<-openxlsx::read.xlsx("../2_Output/2_Methyl/HF.v.NF/DMR/HF.v.NF_Annotated_DiffMeth.xlsx", sheet = "P < 0.05")
colData<-read.csv("../2_Input/_Patient/Index_mep.csv")
colData<-dplyr::filter(colData, Response %in% RESPO, Timing %in% TIM, Pairing %in% PAIRS)
#Filter results by the gene vector
DEGs<-dplyr::filter(Counts, annot.symbol %in% GENES)
DEGs<-dplyr::select(DEGs, annot.symbol, contains("perc.mC"))
rownames(DEGs)<-make.unique(as.character(DEGs$annot.symbol, sep = "."))
tDEGs<-as.data.frame(t(DEGs))
rownames(tDEGs)<-gsub("perc.mC_", "", rownames(tDEGs))
tDEGs = tDEGs[-1, ]

## convert all genes to numeric (from factors)
asNumeric=function(x){as.numeric(as.character(x))}
factorsNumeric=function(d){modifyList(d, lapply(d[, sapply(d, is.factor)], asNumeric))}
##
tDEGs<-factorsNumeric(tDEGs)
tDEGs$Sample.ID<-rownames(tDEGs)
colData.ex<-dplyr::inner_join(tDEGs, colData)
colData.ex$Timing<-factor(colData.ex$Timing, levels = c("CON", "Pre", "Post"))
colData.ex$Response<-factor(colData.ex$Response, levels = c("CON", "NR", "R"))
colData.ex$Etiology<-factor(colData.ex$Etiology, levels = c("CON", "NICM", "ICM"))
colData.ex<-dplyr::group_by_(colData.ex, VAR1, VAR2) #define groups for statistics
groupsize<-dplyr::tally(colData.ex) #define sample size (for SEM)

#Define function for mean and standard deviation for each group
data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sd(x[[col]], na.rm=TRUE))
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
 return(data_sum)
}
#Create a p-value for every summarized data (relative to CON)
t.test2 <- function(m1,m2,s1,s2,n1,n2,m0=0,equal.variance=FALSE)
{
    if( equal.variance==FALSE ) 
    {
        se <- sqrt( (s1^2/n1) + (s2^2/n2) )
        # welch-satterthwaite df
        df <- ( (s1^2/n1 + s2^2/n2)^2 )/( (s1^2/n1)^2/(n1-1) + (s2^2/n2)^2/(n2-1) )
    } else
    {
        # pooled standard deviation, scaled by the sample sizes
        se <- sqrt( (1/n1 + 1/n2) * ((n1-1)*s1^2 + (n2-1)*s2^2)/(n1+n2-2) ) 
        df <- n1+n2-2
    }      
    t <- (m1-m2-m0)/se 
    dat <- c(m1-m2, se, t, 2*pt(-abs(t),df))    
    names(dat) <- c("Difference of means", "Std Error", "t", "p-value")
    return(dat) 
}

########################################################
detach("package:dplyr", unload=TRUE)
plotlist = list()
p<-1
for (i in rownames(DEGs)){
  GENE<-as.name(i) #define a "name" for the variable
  ds <- data_summary(colData.ex, varname=i, groupnames=c(VAR1, VAR2))
  ds<-dplyr::left_join(ds, groupsize)
  ds<-dplyr::mutate(ds, upper=ds[,3]+(sd))
  for(j in seq_along(ds[,4])){
    p[j]=round(t.test2(m1=ds[j,3], m2=ds[1,3], s1=ds[j,4], s2=ds[1,4], n1=ds[j,5], n2=ds[1,5], equal.variance = FALSE)[4], 2)
  }
  ds$p_con<-stars.pval(p)
g<-ggplot(ds, aes_string(x=VAR2, y=i, fill = VAR1)) +
    geom_bar(stat="identity", color="black",
           position=position_dodge()) +
    # geom_point(shape="O",stat="identity", mapping = aes_string(x=VAR2, y=i),data = colData.ex, position=position_dodge(0.9), stroke = 2) +
    geom_errorbar(aes_string(ymin=i, ymax="upper"), width=.2,
                 position=position_dodge(.9)) +
    geom_text(aes(label=p_con), position=position_dodge(width=0.9), vjust=-1, size=5) +
    scale_fill_manual(values=c("white", "black", "darkcyan")) +
    ylim(NA, 1.1*max(ds$upper))
g_plot<-g+labs(title=paste0("Expression - ", i), x=VAR2, y = "Normalized Counts")+
   theme_classic2(base_size = 10)
pdf(file=paste0("../2_Output/2_Methyl/Candidates/", i, "_", VAR1, ".and.", VAR2, "_Methylation.pdf"), width = 2, height = 1.5)
print(g_plot)
dev.off()
plotlist[[i]] = g_plot
}

t<-marrangeGrob(grobs = plotlist, legend, nrow=3, ncol=2)
ggsave(paste0("../2_Output/2_Methyl/Candidates/_DEGs_", VAR1, ".v.", VAR2, ".pdf"), t, width = 6, height = 7)
########################################################

colData.exg<-tidyr::gather(colData.ex, "Gene.Symbol", "DMCs", 1:length(DEGs$annot.symbol))
colData.exg$Gene.Symbol<-factor(colData.exg$Gene.Symbol, levels = rownames(DEGs))
ds_timing<-data_summary(colData.exg, varname="DMCs", groupnames=c("Gene.Symbol", VAR2))
groupsize<-colData.ex %>% group_by(Timing) %>% dplyr::tally() #Calculate sample sizes
ds_timing<-dplyr::left_join(ds_timing, groupsize) #add sample size to summary table
ds_timing<-dplyr::mutate(ds_timing, #make upper and lower bounds based on SEM
                         lower=DMCs-(sd/sqrt(n-1)), 
                         upper=DMCs+(sd/sqrt(n-1))) 
g<-ggplot(ds_timing, aes_string(x="Gene.Symbol", y="DMCs", fill = VAR2)) + 
    geom_bar(stat="identity", color="black", position=position_dodge()) +
    geom_errorbar(aes(ymin=DMCs, ymax=upper), width=.2, position=position_dodge(.9)) + 
    labs(y = "Differential Methylation (%)", x = "Gene Name")+
    theme_classic() +
    scale_fill_manual(values=c('white', "black", "darkcyan")) +
    stat_compare_means(mapping = aes_string(x="Gene.Symbol", y="DMCs"), 
                       data = colData.exg, label = "p.signif")
pdf(file="../2_Output/2_Methyl/Candidates/_DMCs_singlebar.pdf", width = 6, height = 2)
print(g)
dev.off()
```

#Supplemental Table: R Session Information

All packages and setting are acquired using the following command: 
```{r settings}
sinfo<-devtools::session_info()
sinfo$platform
sinfo$packages %>% kable( 
                         align="c", 
                         longtable=T, 
                         booktabs=T,
                         caption="Packages and Required Dependencies") %>% 
    kable_styling(latex_options=c("striped", "repeat_header", "condensed"))
```
